Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2017-09-11 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
abstract and discussion reworked in : paper2 discussion reworking2.docx

### version 13

Target : Malaria journal or Parasites & Vectors

@@

_must be 350 words, currently 349_

##### Background
Insecticide resistance threatens effective vector control, especially for mosquitoes and malaria. Insecticide use strategies that have been recommended to manage resistance include mixtures, sequences and rotations. New insecticides are being developed to address resistance and there is an opportunity to develop use strategies that limit the evolution of further resistance in the short term. There is limited recent, accessible modelling work addressing the evolution of resistance under different operational strategies. Previous work concludes that preferred strategies will be situation specific and there appears to be little recent consensus e.g. on the merits of mixtures. There is an opportunity to improve the level of mechanistic understanding within the operational community of how insecticide resistance can be expected to evolve in response to different strategies.

We provide a concise, accessible description of the use of a flexible model to develop a mechanistic picture the evolution of insecticide resistance and how it is likely to respond to potential insecticide use strategies.  The aim is to reach an audience unlikely to read a more detailed modelling paper by assuming little knowledge of models. The model itself, as described here, represents two independent genes coding for resistance to two insecticides. This allows us to represent the use of insecticides in isolation, sequence and mixtures.

##### Results
We use the model to demonstrate the evolution of resistance under different inputs and how this fits with intuitive reasoning about selection pressure. We show that using an insecticide in a mixture, relative to alone, always prompts slower evolution of resistance to that insecticide. However, when the resistance to both insecticides is considered, resistance thresholds may be reached later for a sequence relative to a mixture. Increasing the ability of insecticides to kill susceptible mosquitoes (effectiveness), has the most influence on favouring a mixture over a sequence.

@@

## Discussion

Resistance responds to insecticide use in the model in ways that we can explain mechanistically through the process of selection. This mechanistic explanation can help us develop a more robust understanding of how we would expect resistance to evolve in the field. We start by explaining the evolution of resistance to a single insecticide, which is relatively straightforward, and move on to explaining the effect of two insecticides in sequences and mixtures.

When single insecticide use was investigated (Figs 3-4) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance. Higher values for cost led to slower evolution of (Fig 4B).

These mechanisms, explaining the  of resistance to single insecticide use under different input values, are summarised in Table 2. Referring to figure 1 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of all genotypes, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

#### Two insecticides
Using two insecticides either in sequence or in mixtures differs from the use of a single insecticide. Where it differs   helps us to understand the dynamics of resistance evolution. Our base scenario (Fig 5A) shows curves for resistance to both insecticides in the mixture that are shallower than those for sequential use. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Because resistant individuals are being killed, the selection pressure for that resistance is lower. Thus each insecticide reduces the selection pressure for resistance to the other and could be said to 'protect' the other. This protection means that evolution of resistance to one insecticide is always slower when it is used in a mixture than when used alone. This slower evolution of resistance to each insecticide in the mixture is shown in Figs. 5-9 by the solid lines (mixture) always being shallower than the dotted lines (sequence) for the same insecticide in the same panel.

_TODO bits to incorporate into discussion :_

_add a paragraph, maybe not even in discussion, walking through how the online UI can be used to see the effect of changing params_

_add para see notebook 11/8/17 saying about inputs that effect time-to-r, but don't effect whether mix or seq is preferred_

_todo, somewhere earlier say good thing about effectiveness & exposure is that they can be modified_

_comments from Hilary_
_This would seem to be a critical point to stress - with examples of how this might be acheived._
_relating to how to maintain control & minimise resistance_

_Limitations that haven't been addressed is the possiblility that resistance is poylgenic, and the possibility of cross resistance (e.g a P450 may cause partial resistance to both)._

_Something further to guide readers on how they might determine dominance would also be useful given how crucial this is._

_Can you give some indications of the type of data that should be collected to help paramertise the model?_

>>>>

### version 14

Target : Malaria journal

@@

_must be 350 words, currently 348_

##### Background
Insecticide resistance threatens effective vector control, especially for mosquitoes and malaria. To manage resistance, recommended insecticide use strategies include mixtures, sequences and rotations. New insecticides are being developed and there is an opportunity to develop use strategies that limit the evolution of further resistance in the short term. A 2013 review of modelling and empirical studies of resistance points to the advantages of mixtures. However, there is limited recent, accessible modelling work addressing the evolution of resistance under different operational strategies. We see an opportunity to improve the level of mechanistic understanding within the operational community of how insecticide resistance can be expected to evolve in response to different strategies.

We provide a concise, accessible description of a flexible model of the evolution of insecticide resistance. We use the model to develop a mechanistic picture the evolution of insecticide resistance and how it is likely to respond to potential insecticide use strategies.  The aim is to reach an audience unlikely to read a more detailed modelling paper. The model itself, as described here, represents two independent genes coding for resistance to two insecticides. This allows us to represent the use of insecticides in isolation, sequence and mixtures.

##### Results
We use the model to demonstrate the evolution of resistance under different inputs and how this fits with intuitive reasoning about selection pressure. We show that using an insecticide in a mixture, relative to alone, always prompts slower evolution of resistance to that insecticide. However, when the resistance to both insecticides is considered, resistance thresholds may be reached later for a sequence relative to a mixture. Increasing the ability of insecticides to kill susceptible mosquitoes (effectiveness), has the most influence on favouring a mixture over a sequence because one highly effective insecticide provides more protection to another in a mixture.

@@

## Discussion
Resistance responds to insecticide use in the model in ways that we can explain mechanistically through the process of selection. This mechanistic explanation can help us develop a more robust understanding of how we would expect resistance to evolve in the field. We start by explaining the evolution of resistance to a single insecticide, which is relatively straightforward, and move on to explaining the effect of two insecticides in sequences and mixtures.

When single insecticide use was investigated (Figs 3-4) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance, whereas higher values for cost led to slower evolution of resistance (Fig 4B).

These mechanisms, explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to figure 1 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

#### Two insecticides
Responses to changing input parameters are different when two insecticides are used either in sequence or in mixtures as compared to when used alone. These differences can help us to understand the dynamics of resistance evolution. In each panel of Figures 5-9 the solid lines, indicating that an insecticide is being used at the same time as another in a mixture, are always shallower than the same colour dashed line indicating the insecticide is being used in isolation as part of a sequence. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Because resistant individuals are being killed, the selection pressure for that resistance is lower. Thus each insecticide reduces the selection pressure for resistance to the other and could be said to 'protect' the other. This protection means that evolution of resistance to one insecticide is always slower when it is used in a mixture than when used alone. This protection has been termed multiple intra-generational killing [@Consortium2013] because individuals have the potential to be killed by more than one insecticide.

Whilst this protection ensures that resistance to each insecticide evolves more slowly in a mixture it does not guarantee that a mixture strategy is preferable  to a sequence once the resistance to both insecticides is taken into account. In a mixture, selection pressure is applied by both insecticides at the same time, whereas in a sequence there is no selection for the insecticide not being used. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 5A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture, so the sequence is still favoured because 2x60 is greater than 90.

This advantage of sequence over mixtures can be removed by increasing the effectiveness of one of the insecticides. Increasing the effectiveness of one insecticide (Fig 5B) decreases the total time for the sequence from 120 to 90 and increases that for the mixture from 90 to 115 thus changing the order and favouring the mixture. The mechanism for this change can be seen by comparing the shape of the curves in Figs 5A & B. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide reaches the resistance threshold of 50% it kills fewer individuals that are resistant to the second insecticide, it loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B. In both cases increasing the effectiveness of insecticides increases their protective effect when used in mixtures and favours mixtures over sequences.

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide (the red curve in this case). The more effective insecticide prompts both a) higher selection pressure for its own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide. The more effective insecticide has a faster evolution of resistance than the less effective one but this is not caused by the presence of the less effective one. (Indeed the presence of the less effective one is still slowing the evolution of resistance to the more effective one, as indicated by the red solid curve for use in the mixture being shallower than the red dashed curve for use alone as part of the sequence).

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 5A to 40 generations in Fig 5C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 5B & 5C). This difference, that exposure reduces times-to-resistance for a mixture and effectiveness increases it, was unexpected for us and we are keen to explore it in more detail. The most likely explanation is that increasing exposure increases overall selection pressure and lowers time to resistance, whereas increasing effectiveness increases mutual protection of the insecticides in a mixture and lowers the speed at which resistance evolves.

#### Costs of resistance
Fitness costs of resistance have been widely reported(review in [@Kliot2012]) but it has been suggested that few of these are directly relevant to control failure in the field [@ffrench-Constant2017]. Because of this likely rarity of costs we did not include them in most of our analyses (with the exception of Fig 9).

Costs of resistance favour mixtures in these analyses because the evolution of resistance is slowed more by cost in a mixture than in a sequence (compare the red solid lines for a mixture to the red dotted line for a sequence in Fig 9C and D). It makes sense that the protective effect of the other insecticide in a mixture combines with the cost of resistance to reduce selection pressure more than in a sequence.

The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes becomes greater. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. However, we do not expect such high costs to occur in operational settings.

Declines in resistance frequencies are likely to occur when there are costs of resistance and the insecticide is not being used [@ffrench-Constant2017]. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where the first insecticide becomes operationally useful again. Consequently, a repeated sequential strategy could offer benefits over mixtures not considered here. To quantify these benefits one would need to use a more detailed metric than the time-to-resistance we have used. We quantified the number of generations until resistance thresholds are first reached. Other measures for quantifying resistance over time can be imagined, for example the mean resistance or proportion of generations below a threshold.

#### Consistency with previous work 
The predicted responses of resistance presented here can be put into the context of previous work and recommendations. A recent review [@Consortium2013] found that 8 out of 10 empirical studies and 11 out of 14 modelling studies favoured mixtures over sequences. In the 3 other modelling studies the relative performance of mixtures and sequences were dependent on the values of other inputs as we also have shown.

The review concluded that the advantage of mixtures is greatest when seven other conditions are satisfied [@Consortium2013] : i) rare starting resistance , ii) independent loci (i.e. no cross resistance), iii) high recombination, iv) high susceptible mortality, v) resistance is recessive, vi) similar persistence of insecticides, and vii) some of population remains untreated. Their points iv) and vii) agree with our results that mixtures are favoured by high effectiveness and low exposure. Some of the other conditions, notably i) the starting frequency of resistance and v) the dominance of resistance did not alter whether mixtures or sequences were favoured in our model [@Levick2017].

Earlier modelling work has made sometimes contrasting predictions and these can be explained by the details of their structure and input choices. One model predicted that "the use of mixtures is always more effective in delaying the onset of resistance often by many orders of magnitude"[@Mani1985]. That model assumed that SS genotypes were always killed, RR always survived, and that a constant 10% of mosquitoes escaped the insecticide. This would be represented in our model by a very high 'insecticide effectiveness' of 1, a 'resistance restoration' of 1 and an 'exposure' of 0.9. This result is consistent with our prediction that the very high effectiveness is likely to favour mixtures over sequences (e.g. see Fig 6D and Table 4).

A subsequent model [@Roush1998] lead to the following more pessimistic conclusion about the potential of mixtures : “As a result of incomplete coverage and residue decay, the mortality of susceptible homozygotes is rarely consistently high enough for pesticide mixtures to be effective” [@Roush1998]. This "mortality of susceptible homozygotes" is equivalent to our insecticide effectiveness. Again this points to the importance of effectiveness and the question of whether the effectiveness levels needed to make mixtures better are attainable operationally. Our result showing that decreasing exposure favours mixtures over sequences is also consistent with figure 3 from that paper [@Roush1998].

One aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (RS) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [Bourguet2000][@Barbosa2011] [@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research into how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer.

Experience from agriculture where single interventions have been associated with a rapid development of resistance have led for calls for a more Integrated Vector Management composed of a series of partially effective tools [@Thomas2012]. It is suggested that such an integrated approach is more sustainable and 'evolution-proof' [@Thomas2012]. A similar combining of interventions has been advocated to cover mosquitoes exhibiting different behaviours [@Killeen2017]. Mosquitoes with different behaviours e.g. feeding outdoors and/or on animals are likely to result in lower exposures to insecticides in nets or sprayed on walls. This lower exposure could favour mixtures over sequences as we have indicated. In both cases our modelling approach could be used to investigate implications for the development of resistance.

#### Caveats
There is much evidence of the involvement of multiple genes in resistance mechanisms. Our model, in common with most previous ones, assumes a single ‘resistance’ gene per insecticide. In most cases we believe this is reasonable. Target-site resistance has been shown to be mostly determined by single genes, and control failures in general have mostly been attributed to the effect of single major genes [@Ffrench-Constant2004] [@Ffrench-Constant2013]. Control doses of insecticides in the field are likely to lead to the selection of single major genes whereas the lower doses used to select laboratory strains are more likely to lead to polygenic resistance [@Ffrench-Constant2013]. To represent polygenic resistance a different 'quantitative genetic' approach would be required. A quantitative genetic model would likely lead to a more rapid initial evolution of resistance, a slower middle phase and a continuing increase. If polygenic resistance is shown in future to be more of a problem then these results could be compared to those of quantitative genetic models.

Even if the assumption that single genes are responsible for control failures is correct there is still considerable other uncertainty around the development of insecticide resistance in operational field settings [@Ranson2016]. The model we present here is able to cover much of the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). There is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences.

In addition we provide an on-line user interface to the model (Fig 2) [@South2017a] enabling readers to investigate model behaviour in areas of parameter space we have not shown. The user-interface allows inputs to be set for two scenarios and results to be viewed side by side. For example you can set a higher insecticide effectivenesses in one scenario to see how this favours mixtures over sequences (Fig 2) or increase the exposure levels to see how this favours sequences.

There are other model behaviours that can be changed by modifying input files. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying input files would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately.

In addition to these uncertainties regarding resistance it is important to remember that our model examines solely the evolution of resistance and not the control of mosquitoes or malaria. Different insecticide use strategies, as well as affecting the evolution of resistance, will also influence both the numbers of vectors and their propensity to transmit the disease [@Rivero2010]. There are also potential knock-on effects of resistance itself; reduced vector transmission of disease (e.g. by reduced lifespan of the vector) or increased transmission (e.g. by increasing the susceptibility  of the vector to the disease) [@Rivero2010]. This model can explore when a mixture strategy may be favoured over a sequence purely for the evolution of resistance. Other operational factors also need to be considered. On the positive side a mixture is likely to kill more vectors and may also limit their ability to transmit the disease. On the negative side a mixture may be more expensive. Other attributes of insecticides such as their repellency are also relevant. Insecticides with some repellency will reduce exposure to the insecticide and thus, as shown here, are likely to slow the evolution of resistance and favour mixtures over sequences. However the reduced exposure resulting from repellency also reduces the population protection provided by the insecticide [@Birget2015].

Our model would predict that the best strategy for limiting the development of resistance in the short term would be to use no insecticide. Of course, using no insecticide is likely to have serious consequences for mosquito and malaria abundance. The challenge is to develop good strategies for delivering mosquito and malaria control in the short term while sustaining their operational capacity in the longer term by restricting the development of resistance.

#### Conclusions
We have developed a flexible, modern model of the evolution of insecticide resistance that can that help us understand the mechanisms by which different insecticide application strategies may be favoured. It replicates previous modelling work and has the potential to investigate further operational strategies for established and new insecticides. For assessing the relative benefits of mixtures and sequences we have demonstrated that mixtures are likely to be favoured by high insecticide effectiveness, low exposure and high costs of resistance. Similar processes will occur in other strategies such as rotations and mosaics. This accessible modelling approach can help to promote discussion and exploration of the likely effect on resistance evolution of different insecticide use strategies in tandem with the collection of empirical data to refine and test predictions.
